---
title: "Review of General Linear Models - Week 2"
format: html
editor: source
---

**Recall the [Jackson Heart Study](https://www.jacksonheartstudy.org/) data from lecture. You can find data from clinic visits 1, 2, and 3 on Canvas.**

#### 1a. Import the data from visit 1. Keep only subjid, sbp, age, HSgrad, diab3cat, and BMI3cat. Make sure that you retain only the complete cases. (Hint: research how to exclude observations with any missing responses.)

```{r, echo = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
library(haven)
data <- read_sas("/Volumes/GoogleDrive/My Drive/STA6257/datasets/analysis1.sas7bdat") %>%
  select(subjid, sbp, age, HSgrad, diab3cat, BMI3cat) %>% 
  na.omit()
```

#### 1b. Create dummy variables for diabetic status (diab3cat) and health status based on body mass index (BMI3cat).

```{r, echo = TRUE}
library(fastDummies)
data <- dummy_cols(data, select_columns = c("diab3cat", "BMI3cat"))
```

#### 2. Model systolic blood pressure at V1 (sbp) as a function of age (years), education (HSgrad; 1=yes, 0=no; use 0 as the reference group), diabetic status (diab3cat; 0=non-diabetic, 1=pre-diabetic, 2=diabetic; use 0 as the reference group), health status based on body mass index (BMI3cat; 0=poor health, 1=intermediate health, 2=ideal health), and all possible two-way interactions. 

```{r, echo = TRUE}
m1 <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
           age:HSgrad + age:diab3cat_1 + age:diab3cat_2 + age:BMI3cat_1 + age:BMI3cat_2 +
           HSgrad:diab3cat_1 + HSgrad:diab3cat_2 + HSgrad:BMI3cat_1 + HSgrad:BMI3cat_2 +
            diab3cat_1:BMI3cat_1 + diab3cat_1:BMI3cat_2 +
            diab3cat_2:BMI3cat_1 + diab3cat_2:BMI3cat_2, data=data)
summary(m1)
```

#### 3. Use the appropriate (single!) hypothesis test to show that we cannot drop all of the interactions **at the same time**.

```{r, echo = TRUE}
m2 <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2, data=data)
anova(m2, m1)
```

#### 4. Use the appropriate hypothesis test to determine if we can drop the interaction between diabetic status and health status based on body mass index. 

```{r, echo = TRUE}
full <- m1
reduced <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
           age:HSgrad + age:diab3cat_1 + age:diab3cat_2 + age:BMI3cat_1 + age:BMI3cat_2 +
           HSgrad:diab3cat_1 + HSgrad:diab3cat_2 + HSgrad:BMI3cat_1 + HSgrad:BMI3cat_2, data=data)
anova(reduced, full)
```

#### 5. Use the appropriate hypothesis test to determine if we can drop the interaction between diabetic status and education. 

```{r, echo = TRUE}
full <- m1
reduced <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
                age:HSgrad + age:diab3cat_1 + age:diab3cat_2 + age:BMI3cat_1 + age:BMI3cat_2 +
                HSgrad:BMI3cat_1 + HSgrad:BMI3cat_2 +
                diab3cat_1:BMI3cat_1 + diab3cat_1:BMI3cat_2 +
                diab3cat_2:BMI3cat_1 + diab3cat_2:BMI3cat_2, 
           data=data)
anova(reduced, full)
```

#### 6. Use the appropriate hypothesis test to determine if we can drop the interaction between health status based on body mass index and education. 

```{r, echo = TRUE}
full <- m1
reduced <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
                age:HSgrad + age:diab3cat_1 + age:diab3cat_2 + age:BMI3cat_1 + age:BMI3cat_2 +
                HSgrad:diab3cat_1 + HSgrad:diab3cat_2 +
                diab3cat_1:BMI3cat_1 + diab3cat_1:BMI3cat_2 +
                diab3cat_2:BMI3cat_1 + diab3cat_2:BMI3cat_2, 
           data=data)
anova(reduced, full)
```

#### 7. Use the appropriate hypothesis test to determine if we can drop the interaction between age and health status based on body mass index. 

```{r, echo = TRUE}
full <- m1
reduced <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
                age:HSgrad + age:diab3cat_1 + age:diab3cat_2 + 
                HSgrad:diab3cat_1 + HSgrad:diab3cat_2 + HSgrad:BMI3cat_1 + HSgrad:BMI3cat_2 +
                diab3cat_1:BMI3cat_1 + diab3cat_1:BMI3cat_2 +
                diab3cat_2:BMI3cat_1 + diab3cat_2:BMI3cat_2, 
           data=data)
anova(reduced, full)
```

#### 8. Use the appropriate hypothesis test to determine if we can drop the interaction between diabetic status and age. 

```{r, echo = TRUE}
full <- m1
reduced <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
                age:HSgrad + age:BMI3cat_1 + age:BMI3cat_2 +
                HSgrad:diab3cat_1 + HSgrad:diab3cat_2 + HSgrad:BMI3cat_1 + HSgrad:BMI3cat_2 +
                diab3cat_1:BMI3cat_1 + diab3cat_1:BMI3cat_2 +
                diab3cat_2:BMI3cat_1 + diab3cat_2:BMI3cat_2, 
           data=data)
anova(reduced, full)
```

#### 9. Use the appropriate hypothesis test to determine if we can drop the interaction between age and education. 

```{r, echo = TRUE}
full <- m1
reduced <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
                age:diab3cat_1 + age:diab3cat_2 + age:BMI3cat_1 + age:BMI3cat_2 +
                HSgrad:diab3cat_1 + HSgrad:diab3cat_2 + HSgrad:BMI3cat_1 + HSgrad:BMI3cat_2 +
                diab3cat_1:BMI3cat_1 + diab3cat_1:BMI3cat_2 +
                diab3cat_2:BMI3cat_1 + diab3cat_2:BMI3cat_2, 
           data=data)
anova(reduced, full)
```

#### 10. Construct a "final model" that does not have the interactions you deemed insignificant in questions 4 through 9. 

```{r, echo = TRUE}
m2 <- lm(sbp ~ age + HSgrad + diab3cat_1 + diab3cat_2 + BMI3cat_1 + BMI3cat_2 +
                age:diab3cat_1 + age:diab3cat_2 + age:BMI3cat_1 + age:BMI3cat_2 +
                HSgrad:BMI3cat_1 + HSgrad:BMI3cat_2, 
           data=data)
summary(m2)
```

#### 11. Use the model from question 10 to find the model for a 50 year old high school graduate.

\begin{align*}
\hat{y} =& `r 106.25710+0.35979*50+0.50995*1` - `r 0.42300+0.03161*50` \times \text{ pre-diabetic} + `r 16.72458-0.25546*50` \times \text{diabetic} \\ &- `r 4.29887+0.13274*50-5.52355` \times \text{ BMI}_{\text{moderate}} - `r 13.85030+0.21018*50+0.13121` \times \text{ BMI}_{\text{poor}}
\end{align*}

<!--- 
Note: I am having R do the calculation for me with the stuff in backticks. 
      When I plugged in values, I put them where they would go after plugging in.
      (Algebraically, combining like terms.)
      e.g., the coefficient for age is multiplied by 50, which is just a scalar, 
      so it's added to the intercept. 
--->

#### 12. Use the model from question 11 to interpret the slopes corresponding to diabetic status and health status as defined by body mass index.

- On average, the SBP of 50 year old pre-diabetics that have at least a high school education is 2 mmHg lower than that of non-diabetics.

- On average, the SBP of 50 year old diabetics that have at least a high school education is 4 mmHg higher than that of non-diabetics.

- On average, the SBP of 50 year olds in moderate health, as defined by BMI, and have at least a high school education is 5.4 mmHg lower than those in poor health as defined by BMI.

- On average, the SBP of 50 year olds in ideal health, as defined by BMI, and have at least a high school education is 24.5 mmHg lower than those in poor health as defined by BMI. 

#### 13. Use the model from question 10 to find the model for a pre-diabetic person that is in poor health.

\begin{align*}
\hat{y} =& `r 106.25710- 0.42300` + 0.35979 \times \text{ age} + 0.50995 \times \text{ HS grad} 
\end{align*}

#### 14. Use the model from question 13 to interpret the slope corresponding to age. 

- On average, for pre-diabetics in poor health, as defined by BMI, for a 1 year increase in age, we expect SBP to increase by 0.36 mmHg.

- On average, for pre-diabetics in poor health, as defined by BMI, for a 5 year increase in age, we expect SBP to increase by `r 0.36*5` mmHg.

- On average, for pre-diabetics in poor health, as defined by BMI, for a 10 year increase in age, we expect SBP to increase by `r 0.36*10` mmHg.

#### 15. (Challenge!) Construct a visualization to help with interpreting this model. Systolic blood pressure should be on the *y*-axis, age should be on the *x*-axis. You choose what lines to create. 





