---
title: "Review of General Linear Models - Week 2"
format: html
editor: source
---

**Recall the [Jackson Heart Study](https://www.jacksonheartstudy.org/) data from lecture. You can find data from clinic visits 1, 2, and 3 on Canvas.**

#### 1a. Import the data from visit 1. Keep only subjid, sbp, age, HSgrad, diab3cat, and BMI3cat. Make sure that you retain only the complete cases. (Hint: research how to exclude observations with any missing responses.)

```{r, echo = TRUE, message = FALSE}

```

#### 1b. Create dummy variables for diabetic status (diab3cat) and health status based on body mass index (BMI3cat).

```{r, echo = TRUE}

```

#### 2. Model systolic blood pressure at V1 (sbp) as a function of age (years), education (HSgrad; 1=yes, 0=no; use 0 as the reference group), diabetic status (diab3cat; 0=non-diabetic, 1=pre-diabetic, 2=diabetic; use 0 as the reference group), health status based on body mass index (BMI3cat; 0=poor health, 1=intermediate health, 2=ideal health), and all possible two-way interactions. 

```{r, echo = TRUE}

```

#### 3. Use the appropriate (single!) hypothesis test to show that we cannot drop all of the interactions **at the same time**.

```{r, echo = TRUE}

```

#### 4. Use the appropriate hypothesis test to determine if we can drop the interaction between diabetic status and health status based on body mass index. 

```{r, echo = TRUE}

```

#### 5. Use the appropriate hypothesis test to determine if we can drop the interaction between diabetic status and education. 

```{r, echo = TRUE}

```

#### 6. Use the appropriate hypothesis test to determine if we can drop the interaction between health status based on body mass index and education. 

```{r, echo = TRUE}

```

#### 7. Use the appropriate hypothesis test to determine if we can drop the interaction between age and health status based on body mass index. 

```{r, echo = TRUE}

```

#### 8. Use the appropriate hypothesis test to determine if we can drop the interaction between diabetic status and age. 

```{r, echo = TRUE}

```

#### 9. Use the appropriate hypothesis test to determine if we can drop the interaction between age and education. 

```{r, echo = TRUE}

```

#### 10. Construct a "final model" that does not have the interactions you deemed insignificant in questions 4 through 9. 

```{r, echo = TRUE}

```

#### 11. Use the model from question 10 to find the model for a 50 year old high school graduate.

#### 12. Use the model from question 11 to interpret the slopes corresponding to diabetic status and health status as defined by body mass index.

#### 13. Use the model from question 10 to find the model for a pre-diabetic person that is in poor health.

#### 14. Use the model from question 13 to interpret the slope corresponding to age. 

#### 15. (Challenge!) Construct a visualization to help with interpreting this model. Systolic blood pressure should be on the *y*-axis, age should be on the *x*-axis. You choose what lines to create. 

```{r, echo = TRUE}

```




