---
title: "Generalized Linear Models & Gamma Regression - Week 3"
format: html
self-contained: true
editor: source
---

**Recall the [Jackson Heart Study](https://www.jacksonheartstudy.org/) data from lecture. You can find data from clinic visits 1, 2, and 3 on Canvas. For any question requiring an $\alpha$, assume $\alpha=0.05$.**

#### 1a. Import the data from visit 1. Keep only subjid, HbA1c, age, totchol, HTN, HSgrad, and BMI3cat. Retain only the complete cases. 

```{r, echo = TRUE, message = FALSE}

```

#### 1b. Create dummy variables for health status based on body mass index (*BMI3cat*).

```{r, echo = TRUE}

```

#### 2. Model hemoglobin A1c at V1 (*HbA1c*) as a function of age (*age*; years), total cholesterol (*totchol*; mg/dL), hypertension status(*HTN*; 1=yes, 0=no; use 0 as the reference group), education (*HSgrad*; 1=yes, 0=no; use 0 as the reference group), and health status based on body mass index (*BMI3cat*'s dummy variables; 0=poor health, 1=intermediate health, 2=ideal health; use 0 as the reference group), and all possible two-way interactions involving hypertension status. 

```{r, echo = TRUE}

```

#### 3. Use the appropriate (single!) hypothesis test to show that we cannot drop all of the interactions **at the same time**.

```{r, echo = TRUE}

```

#### 4. Use the appropriate hypothesis test to determine if we can drop the interaction between hypertension status and health status as defined by body mass index. 

```{r, echo = TRUE}

```

#### 5. Use the appropriate hypothesis test to determine if we can drop the interaction between hypertension status and education. 

```{r, echo = TRUE}

```

#### 6. Use the appropriate hypothesis test to determine if we can drop the interaction between hypertension status and total cholesterol. 

```{r, echo = TRUE}

```

#### 7. Use the appropriate hypothesis test to determine if we can drop the interaction between hyprertension status and age. 

```{r, echo = TRUE}

```

#### 8. Construct a "final model" that does not have the interactions you deemed insignificant in questions 4 through 7. 

```{r, echo = TRUE}

```

#### 9. Use the appropriate hypothesis tests to determine which main effects are significant predictors of hemoglobin A1c. (Reminder: we cannot discuss main effects when they are involved in interaction terms.)

```{r, echo = TRUE}

```

#### 10. Use the model from question 8 to find the individual models for those with and without hypertension.

#### 11. Provide interpretations of the slopes in the model for those with hypertension.

#### 12. Provide interpretations of the slopes in the model for those without hypertension.

#### 13. (Challenge!) Construct a visualization to help with interpreting this model. Hemoglobin A1c should be on the *y*-axis, age should be on the *x*-axis. You choose what lines to create. 

```{r, echo = TRUE}

```




